### 2014-11-01 implementation.tex
implementation

In this section, we describe the instantiation of \name on top of off-the-shelf
eventually consistent distributed store Cassandra. We particularly focus on the
challenges in realizing an efficient implementation of our model.

@@

\name is implemented as a shallow extension of Haskell on the Glasgow Haskell
Compiler (GHC). We utilize Template Haskell functionality to implement static
contract classification. Our proof obligations are discharged automatically
with the help of Z3~\cite{} SMT solver. Figure~\ref{fig:impl_mod} describes the
overall system architecture. The main idea in our implementation of \name is to
offload the majority of concerns related to data management such as
replication, fault tolerance, availability and convergence to Cassandra, an
off-the-shelf eventually consistent store (called \emph{backing store} in the
sequel).

Replicated data type implementation and the stronger consistency semantics are
enforced in the \emph{shim layer}. Our implementation supports eventual, causal
and strong consistency for operations, and RC, MAV and RR transaction
semantics. All of this functionality is implemented on top of the standard
interface exposed by Cassandra, and illustrates the portability of \name
programming model. Complex business logic is implemented with the help of
transactions on the web-server side, with clients interacting with the
web-service over REST interface. From an engineering perspective, resorting to
an off-the-shelf backing store meant that our entire implementation comprised of
only around 2500 lines of Haskell code, which is packaged as a Haskell
library\footnote{url to the hackage}.

@@

The shim layer maintains a causally consistent in-memory snapshot of a subset
of objects in the backing store, by explicitly tracking dependencies introduced
between the effects due to visibility, session and same transaction relations.
The dependence tracking is similar to the techniques presented in~\cite{BoltOn}
and~\cite{Eiger}, with the usual optimizations making use of transitivity
properties for minimizing the number of dependencies. Shim layer performs the
reductions associated with replicated datatype operations corresponding to
client requests. As the backing store provides durability, convergence and
fault tolerance, each shim layer node simply acts as a soft-state cache, and
can safely be terminated at any instant. Similarly, more shim layer nodes can
be spawned on demand.

@@

Every effect generated as a result of an effectful operation on an object
inserts a new row $(o,e,vis,txn,val)$ into the backing store, where $o$ and $e$
are object and (unique) effect identifiers, $vis$ is the set of identifiers of
effects visible to this operation, $txn$ is an optional transaction identifier,
and $val$ is the value associated with the effect (eg: \cf{Withdraw 50}). The
shim layer periodically fetches updates from the backing store for those objects
which were accessed since updates were last fetched. Since causally consistent
operations require an up-to-date view of the current session, the shim layer
node synchronously fetches operations if the causally preceding operations in
the current session are not available in the cache.  Strongly consistent
operations are performed after obtaining exclusive leases on objects. The lease
mechanism is implemented with the help of Cassandra's support for conditional
updates and expiring columns.

@@

While Cassandra offers all-or-nothing failure semantics for multiple writes
through batching, readers may witness the initial write while the batch is in
progress. \name implements atomic visibility by exploiting shim layer causality
guarantee -- an effect is included only if all the effects if depends on are
also included.

For every transaction, we instantiate a special transaction marker effect $m$,
but importantly, do not insert into the backing store. $m$ is included as a
dependence to every effect (let $e$ be one such effect) generated in the
transaction. Since $m$ has not yet been written to the store, no operation will
witness $e$ while the transaction in progress. Let $S$ be the set of effects
generated in the transaction. At the end of the transaction, we insert $m$ into
the backing store with $S$ as its dependence. Now any replica which includes
$e$, must include $m$, and transitively must include every effect from the
transaction. This ensures atomicity and satisfies the RC requirement.

@@

The main challenge in realizing an efficient implementation of operation-based
replicated data types is that the state of the object i.e., the set of effects
grows with every effectful operation on the object. If left unchecked, the
operations slow down over time, until the shim layer memory or backing store
disk runs out of memory. Luckily, the state of the operation-based replicated
data type can often be summarized to an \emph{observably equivalent} smaller
state. For example,

@@

Since the semantics of summarization depends on the semantics of the data type,
we expect the programmer to provide a summarization function for each RDT with
the following type:

\begin{codehaskell}
summarize :: Effect e => [e] -> [e]
\end{codehaskell}

\noindent with the intention that the length of the result is smaller that the
length of the argument. We utilize the \cf{summarize} function to summarize the
object state both in the shim layer node and the backing store, typically when
the number of effects on an object crosses a tunable threshold. Shim layer
summarization is straight-forward; a summarization thread takes the local lock
on the object, and replaces its state with the summarized state. The shim layer
node remains unavailable for that particular object during summarization
(usually a few milliseconds).

Compared to the shim layer, summarization in the backing store is more
complicated. The main challenge is that unlike the shim layer, summarization
cannot run as an atomic operation. Summarization in the backing store involves
deleting previously inserted rows and inserting new rows, where each row
corresponds to an effect. It is essential that concurrent client operations are
permitted, but are not allowed to witness the intermediate state of the
summarization process. To this end, we adopt a summarization strategy that
builds on the causality property of the store. The details of the summarization
are beyond the scope of the paper, and is presented in detail in the tech
report~\cite{}.

>>>>

\name is implemented as a shallow extension of GHC Haskell and runs on top
of Cassandra, an off-the-shelf eventually consistent distributed data (or
backing) store responsible for all data management issues (i.e.,
replication, fault tolerance, availability, and convergence).  Template
Haskell is used implement static contract classification, and proof
obligations are discharged with the help of the Z3~\cite{} SMT
solver. Figure~\ref{fig:impl_mod} describes the overall system architecture.

@@

Replicated data types and the stronger consistency semantics are
implemmented and enforced in the \emph{shim layer}. Our implementation
supports eventual, causal, and strong consistency for data type operations,
and RC, MAV, and RR semantics for transactions.  This functionality is
implemented entirely on top of the standard interface exposed by Cassandra.
From an engineering perspective, leveraging an off-the-shelf data store
enables an implementation comprising roughly only 2500 lines of Haskell
code, packaged as a library\footnote{url to the hackage}.

%% Complex business logic is implemented with the help of
%% transactions on the web-server side, with clients interacting with the
%% web-service over REST interface.

@@

The shim layer maintains a causally consistent in-memory snapshot of a
subset of objects in the backing store, by explicitly tracking dependencies
introduced between effects due to visibility, session and same transaction
relations.  Dependence tracking is similar to the techniques presented
in~\cite{BoltOn} and~\cite{Eiger}. Because Cassandra provides durability,
convergence and fault tolerance, each shim layer node simply acts as a
soft-state cache, and can safely be terminated at any point. Similarly, new
shim layer nodes can be spawned on demand.

@@

Every generated effect on an object inserts a new row
$(o,e,\mathit{vis},\mathit{txn},\mathit{val})$ into the backing store, where
$o$ and $e$ are object and (unique) effect identifiers, \emph{vis} is the
set of effect identifiers visible to this operation, \emph{txn} is an
optional transaction identifier, and \mathi{val is the value associated with the
effect. (eg: \cf{Withdraw 50}).  Since causally consistent operations require an
up-to-date view of the current session, the shim layer node synchronously
fetches operations if the causally preceding operations in the current
session are not available in its cache.  Strongly consistent operations are
performed after obtaining exclusive leases on objects. The lease mechanism
is implemented with the help of Cassandra's support for conditional updates
and expiring columns.

@@

While Cassandra offers all-or-nothing failure semantics for multiple writes
through batching, readers may witness the initial write while the batch is
in progress. \name implements atomic visibility by exploiting the shim layer
causality guarantee -- an effect is included only if all the effects if
depends on are also included.

For every transaction, we instantiate a special transaction marker effect
$m$, but importantly, do not insert this into the backing store. $m$ is
included as a dependence to every effect (let $e$ be one such effect)
generated in the transaction. Since $m$ has not yet been written to the
store, no operation will witness $e$ while the transaction in progress. Let
$S$ be the set of effects generated in the transaction. At the end of the
transaction, we insert $m$ into the backing store with $S$ as its
dependence. Now any replica which includes $e$, must include $m$, and
transitively must include every effect from the transaction. This ensures
atomicity and satisfies the RC requirement.

@@

The main challenge in realizing an efficient implementation of
operation-based replicated data types is that the state of the object
defined in terms of its set of effects, grows with every effectful
operation. If left unchecked, performance slow down over time, until the
shim layer memory or backing store disk runs out of memory. Luckily, the
state of the operation-based replicated data type can often be summarized to
an \emph{observably equivalent} smaller state. For example,

@@

\noindent Since the semantics of summarization depends on the semantics of the data
type, our programming model provides a summarization function for each RDT:
\begin{codehaskell}
summarize :: Effect e => [e] -> [e]
\end{codehaskell}
\noindent with the intention that the length of the result is smaller that the
length of the argument. We utilize the \cf{summarize} function to summarize the
object state both in the shim layer node and the backing store, typically when
the number of effects on an object crosses a tunable threshold. Shim layer
summarization is straight-forward; a summarization thread takes the local lock
on the object, and replaces its state with the summarized state. The shim layer
node remains unavailable for that particular object during summarization
(usually a few milliseconds).

Summarization in the backing store is more complicated. The main challenge
is that unlike the shim layer, summarization cannot run as an atomic
operation. Summarization in the backing store involves deleting previously
inserted rows and inserting new rows, where each row corresponds to an
effect. To ensure high availability, it is essential that concurrent client
operations are permitted, but are not allowed to witness the intermediate
state of the summarization process. To this end, we adopt a summarization
strategy that builds on the causality property of the store. The details of
the summarization are beyond the scope of the paper, and is presented in
detail in the tech report~\cite{}.
